---
title: How do socioeconomic and demographic factors influence the prevalence of different types of disabilities across U.S. states?
author: "Jingwen Zhong"
output: html_document
---
link to GitHub repository: https://github.com/lisazhong13/JSC370-finalproject
downloadable HTML link: https://github.com/lisazhong13/JSC370-finalproject/JSC370-Midterm.html
```{r include=FALSE}
# load all neccessary libraries
library(httr)
library(readr)
library(dplyr)
library(rvest)
library(jsonlite)
library(tidyr)
library(ggplot2)
library(ggcorrplot)
library(stats)
library(cluster)
library(knitr)
library(kableExtra)
library(tibble)
library(sf)
library(tmap)
library(tidyverse)
library(tigris)
library(leaflet)
library(htmltools)
```

### Introduction

Disabilities affect millions of individuals across the United States, but their prevalence varies significantly from state to state. These variations may be influenced by a range of socioeconomic factors, including income levels, education attainment, and employment status, as well as demographic factors such as age and race. This project aims to explore the question: "How do socioeconomic indicators and demographic characteristics impact the prevalence of different types of disabilities across U.S. states?"

To address this question, I utilize publicly available data from multiple sources, integrating information on disability prevalence with key economic and demographic variables. By examining these relationships, this study seeks to provide valuable insights into the underlying factors driving disparities in disability rates. Such findings can help inform policy decisions, guide resource allocation, and enhance accessibility and support services for individuals with disabilities across different communities.

I hypothesize that U.S. states with higher poverty rates, lower levels of healthcare coverage, and greater economic inequality will exhibit a higher prevalence of disabilities. Specifically, I expect Medicaid and Medicare spending to be strongly associated with reported disability rates, while employer-sponsored insurance coverage will be negatively correlated with disability prevalence.

### Methods

In this project, the primary dataset is from Centers for Disease Control and Prevention (CDC) the Disability and Health Data System (DHDS) using API. Since the API has a 1000-row limit, data was retrieved in chunks using limit and offset parameters, and loaded into R by the httr and readr libraries. The DHDS data contains the prevalence and disability population in 2021 and 2022 in different states in the U.S.. To better investigate my research question, I complemented the dataset with socioeconomic indicators and demographic factors by scraping uninsured rate data from American Community Survey (ACS) and using the U.S. Census Bureau API, and poverty data, healthcare access data like health insurance coverage data, hospital number data, and medicaid spending data, as well as additional disability percentage data, all downloaded from Kaiser Family Foundation (KFF) website. I merged those datasets together based on the state names, some locations in the primary dataset like "HHS Region 1-10", "Puerto Rico", "Guam", "U.S. Virgin Islands", "United States, DC & Territories" were excluded.

Since many socioeconomic data were missing in 2022, I focused mainly on analyzing data from 2021 and filtered out those from 2022. Then, I renamed the variables to make them easier to understand and changed some character variables into numerical ones. After this, I handled missing values by dropping them. In the end, my cleaned dataset had 378 entries and 30 variables.


```{r include=FALSE}
# Base API endpoint
base_url <- "https://data.cdc.gov/resource/s2qv-b27b.csv"

# Function to fetch data with pagination
fetch_data <- function(offset) {
  url <- paste0(base_url, "?$limit=1000&$offset=", offset)
  temp_file <- tempfile(fileext = ".csv")  # Temporary file to store data
  GET(url, write_disk(temp_file, overwrite = TRUE))  # Fetch data and save
  read_csv(temp_file)  # Read CSV into dataframe
}

# Fetch first 1000 rows
df1 <- fetch_data(0)  

# Fetch next 1000 rows
df2 <- fetch_data(1000)  

# Fetch next 1000 rows
df3 <- fetch_data(2000)  

# Fetch remaining rows (592 rows)
df4 <- fetch_data(3000)  

# Combine all dataframes
df_cdc <- bind_rows(df1, df2, df3, df4)

# Preview merged data
df_cdc
```
```{r include=FALSE}
# Select and rename columns
df_cdc_cleaned <- df_cdc %>% filter(year == 2021) %>%
  select(Year = year, Location = locationdesc, Disability = response, 
         Prevalence = data_value, DisabilityPopulation = weightednumber)

# Preview the updated dataframe
df_cdc_cleaned
```
```{r include=FALSE}
# Define API endpoint for uninsured rate (ACS 2021, Table B27010)
census2021_url <- "https://api.census.gov/data/2021/acs/acs1?get=NAME,B27010_001E,B27010_017E&for=state:*"

# Fetch data
response <- GET(census2021_url)
data <- content(response, as = "text")
json_data <- fromJSON(data)

# Convert to DataFrame
df_census_2021 <- as.data.frame(json_data[-1,], stringsAsFactors = FALSE)
colnames(df_census_2021) <- json_data[1,]  # Set column names

# Rename columns for clarity
df_census_2021 <- df_census_2021 %>%
  rename(
    State = NAME,
    Total_Population = B27010_001E,
    Uninsured_Population = B27010_017E
  ) %>%
  mutate(
    Total_Population = as.numeric(Total_Population),
    Uninsured_Population = as.numeric(Uninsured_Population),
    Uninsured_Rate = (Uninsured_Population / Total_Population) * 100
  ) %>%
  select(State, Uninsured_Rate)

# Preview data
df_census_2021
```
```{r include=FALSE}
# Standardize state names in df_cdc_cleaned
df_cdc_cleaned <- df_cdc_cleaned %>%
  mutate(Location = tolower(Location))

# Standardize state names in df_census_2021 and df_census_2022
df_census_2021 <- df_census_2021 %>%
  mutate(State = tolower(State))

# Merge 2021 data with 2021 uninsured rate
df_merged <- df_cdc_cleaned %>%
  inner_join(df_census_2021, by = c("Location" = "State"))

# Preview final merged dataset
df_merged
```
```{r include=FALSE}
poverty <- read.csv("data/Poverty_by_Age.csv")
insurance <- read.csv("data/Health_Insurance_Coverage_of_the_Total_Population.csv")
disability_percentage <- read.csv("data/Disability_Percentage.csv")
hospital <- read.csv("data/Hospitals.csv")
medicaid <- read.csv("data/Medicaid_Spending.csv")
poverty
insurance
disability_percentage
hospital
medicaid
```
```{r include=FALSE}
poverty <- poverty %>%
  mutate(Location = tolower(Location))
insurance <- insurance %>%
  mutate(Location = tolower(Location))
hospital <- hospital %>%
  mutate(Location = tolower(Location))
disability_percentage <- disability_percentage %>%
  mutate(Location = tolower(Location))
medicaid <- medicaid %>%
  mutate(Location = tolower(Location))

# Start by merging df_merged with the first dataset
df_final <- df_merged %>%
  inner_join(poverty, by = "Location") %>%
  inner_join(insurance, by = "Location") %>%
  inner_join(disability_percentage, by = "Location") %>%
  inner_join(hospital, by = "Location") %>%
  inner_join(medicaid, by = "Location")

df_final <- df_final %>% drop_na()
# Preview the final merged dataset
df_final
```
```{r include=FALSE}
# Convert relevant columns to numeric
df_final <- df_final %>%
  mutate(
    Military = as.numeric(Military),
    Total.Medicare.Part.A.Enrollees = as.numeric(gsub(",", "", Total.Medicare.Part.A.Enrollees)),
    Total.Hospitals = as.numeric(Total.Hospitals),
    Total.Hospital.Beds = as.numeric(gsub(",", "", Total.Hospital.Beds)),
    Short.Stay.Hospitals = as.numeric(Short.Stay.Hospitals),
    Short.Stay.Hospital.Beds = as.numeric(gsub(",", "", Short.Stay.Hospital.Beds)),
    Critical.Access.Hospitals = as.numeric(Critical.Access.Hospitals),
    Critical.Access.Hospital.Beds = as.numeric(gsub(",", "", Critical.Access.Hospital.Beds)),
    All.Other.Hospitals = as.numeric(All.Other.Hospitals),
    All.Other.Hospital.Beds = as.numeric(gsub(",", "", All.Other.Hospital.Beds)),
    Seniors = as.numeric(gsub("[$,]", "", Seniors)),  # Remove $ and ,
    Individuals.with.Disabilities = as.numeric(gsub("[$,]", "", Individuals.with.Disabilities)),
    Adult = as.numeric(gsub("[$,]", "", Adult)),
    Children = as.numeric(gsub("[$,]", "", Children)),
    Newly.Eligible.Adult = as.numeric(gsub("[$,]", "", Newly.Eligible.Adult)),
    Total.y = as.numeric(gsub("[$,]", "", Total.y))
  ) %>%
select(
    Year, 
    State = Location, 
    Disability, 
    DisabilityPrevalence = Prevalence, 
    DisabilityPopulation, 
    UninsuredRate = Uninsured_Rate, 
    PovertyChildren = Children.0.18, 
    PovertyAdult = Adults.19.64, 
    PovertySenior = X65., 
    PovertyTotal = Total.x, 
    EmployerInsuranceCoverage = Employer, 
    MedicaidInsuranceCoverage = Medicaid, 
    MedicareInsuranceCoverage = Medicare, 
    MilitaryInsuranceCoverage = Military, 
    SelfCareDisability = Self.Care, 
    HearingDisability = Hearing, 
    VisionDisability = Seeing, 
    IndependentLivingDisability = Living.Independently, 
    MobilityDisability = Ambulatory, 
    CognitiveDisability = Cognitive, 
    AnyDisability = Any.Disability, 
    TotalMedicareEnrollees = Total.Medicare.Part.A.Enrollees, 
    Total.Hospitals, 
    Total.Hospital.Beds, 
    TotalHospitalBedsPer1000Enrollees = Total.Hospital.Beds.Per.1.000.Part.A.Enrollees, 
    SeniorMedicaidSpending = Seniors, 
    DisabilityMedicaidSpending = Individuals.with.Disabilities, 
    AdultMedicaidSpending = Adult, 
    ChildrenMedicaidSpending = Children, 
    TotalMedicaidSpending = Total.y
  )
df_final <- df_final %>% drop_na()
```

```{r include=FALSE}
df_final
```

```{r include=FALSE}
ggplot(df_final, aes(x = State, y = DisabilityPrevalence, fill = State)) +
  geom_boxplot() +
  labs(title = "Disability Rates Across U.S. States", y = "Disability Prevalence (%)", x = "State") +
  coord_flip() +
  theme_minimal()
```
### Preliminary Results

With this cleaned dataset, I now proceed to an exploratory analysis, using descriptive statistics and data visualizations to uncover patterns in disability prevalence across U.S. states. The following section presents preliminary findings, including distribution patterns, correlation insights, and statistical tests that provide a foundation for deeper analysis.

```{r echo=FALSE}
# Select relevant columns
df_long <- df_final %>%
  select(State, SelfCareDisability, HearingDisability, VisionDisability, 
         IndependentLivingDisability, MobilityDisability, CognitiveDisability, AnyDisability) %>%
  pivot_longer(cols = -State, names_to = "Disability_Type", values_to = "Proportion")

ggplot(df_long, aes(x = State, y = Proportion, fill = Disability_Type)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Figure 1: Proportion of Each Disability Type in Each State",
       x = "State", y = "Proportion", fill = "Disability Type") +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

From the stacked barcharts above, "Any Disability" dominates across all states. "Cognitive Disability", "Independent Living Disability", and "Mobility Disability" are visibly large across most states, suggesting that both mental and physical disabilities are significant components of disability prevalence. Some states may have a higher proportion of cognitive disabilities, possibly indicating different healthcare access or demographic variations. "Hearirng Disabilities" and "Vision Disabilities" appear in smaller proportions in most states.

```{r echo=FALSE}
# Assuming `df_final` is your dataset
numeric_vars <- df_final %>% select(AnyDisability, PovertyTotal, TotalMedicaidSpending, TotalMedicareEnrollees, DisabilityMedicaidSpending, UninsuredRate)

# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Plot the correlation heatmap with a legend title
ggcorrplot(cor_matrix, 
           lab = TRUE,             # Display correlation values
           hc.order = TRUE,        # Hierarchical clustering order
           type = "lower",         # Show lower triangle
           colors = c("blue", "white", "red"),  # Color gradient
           legend.title = "Correlation") +  # Set the legend title
  ggtitle("Figure 2: Correlation Matrix of Disability & Economic Factors") +  # Add plot title
  theme(plot.title = element_text(hjust = 0.5, size = 14))  # Center title
```

TotalMedicaidSpending and TotalMedicareEnrollees have a very high positive correlation, indicating states with more Medicare enrollees tend to have higher total Medicaid spending. They are also highly correlated with DisabilityMedicaidSpending, suggesting that disability-related Medicaid spending is a large component of total Medicaid spending. AnyDisability and PovertyTotal have a relatively high correlation, meaning that higher poverty levels are associated with a higher percentage of people reporting disabilities. There are also some negatively correlated factors, DisabilityMedicaidSpending and UninsuredRate have a moderate negative correlation, suggesting that as disability-related Medicaid spending increases, the uninsured rate tends to decrease. A similar negative correlation appear to TotalMedicaidSpending and UninsuredRate, implying that increased Medicaid spending is linked to a lower uninsured rate.

```{r include=FALSE}
# Get US state boundaries (excluding territories)
us_states <- states(cb = TRUE, year = 2020) %>%
  filter(!STUSPS %in% c("PR", "VI", "GU", "MP", "AS"))  # Remove territories

us_states <- us_states %>% mutate(NAME = tolower(NAME))
```

```{r warning=FALSE, echo=FALSE}
# Merge your dataset with the state shapefile
map_data <- left_join(us_states, df_final, by = c("NAME" = "State"))  # Ensure matching column names

# Convert `disabilityprevalence` to numeric (in case it's a character column)
map_data$AnyDisability <- as.numeric(map_data$AnyDisability)

# Define a color palette
pal <- colorBin("Reds", domain = map_data$AnyDisability, bins = 5)

# Create an interactive Leaflet map
leaflet(map_data) %>%
  addTiles() %>%  # Add base map
  addPolygons(
    fillColor = ~pal(AnyDisability),
    color = "white",  # Border color
    weight = 1,
    smoothFactor = 0.5,
    opacity = 1,
    fillOpacity = 0.7,
    label = ~paste0(NAME, ": ", AnyDisability, "%"),  # Hover label
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(pal = pal, values = ~AnyDisability, opacity = 0.7,
            title = "Disability Prevalence (%)", position = "bottomright") %>%
    addControl(
    html = HTML("<div style='font-size:18px; font-weight:bold; text-align:center;'>Figure 3: Disability Distribution by Region</div>"), 
    position = "topleft"
  )
```
The map shows disability prevalence (%) across the United States, with different states shaded based on their percentage of the population with disabilities. Higher disability prevalence (darker red shades) is observed in Southern states such as Kentucky, New Mexico, Oklahoma, Arkansas, Louisiana, Alabama. Lower disability prevalence (lighter shades) appears in Western states like California, Utah, and Colorado, and in some parts of the Northeast. Alaska appears to have relatively higher disability prevalence compared to many other states. Disability prevalence is often correlated with poverty levels, healthcare access, and demographics.
```{r include=FALSE}
# Define a color palette
pal <- colorBin("Reds", domain = map_data$PovertyTotal, bins = 5)

# Create an interactive Leaflet map
leaflet(map_data) %>%
  addTiles() %>%  # Add base map
  addPolygons(
    fillColor = ~pal(PovertyTotal),
    color = "white",  # Border color
    weight = 1,
    smoothFactor = 0.5,
    opacity = 1,
    fillOpacity = 0.7,
    label = ~paste0(NAME, ": ", PovertyTotal, "%"),  # Hover label
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(pal = pal, values = ~PovertyTotal, opacity = 0.7,
            title = "Poverty Probability (%)", position = "bottomright")
```

```{r warning=FALSE, echo=FALSE}
# Define a color palette
pal <- colorBin("Reds", domain = map_data$PovertyTotal, bins = 5)

# Create an interactive Leaflet map
leaflet(map_data) %>%
  addTiles() %>%  # Add base map
  addPolygons(
    fillColor = ~pal(PovertyTotal),
    color = "white",  # Border color
    weight = 1,
    smoothFactor = 0.5,
    opacity = 1,
    fillOpacity = 0.7,
    label = ~paste0(NAME, ": ", PovertyTotal, "%"),  # Hover label
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(pal = pal, values = ~PovertyTotal, opacity = 0.7,
            title = "Poverty Probability (%)", position = "bottomright") %>%
  addControl(
    html = HTML("<div style='font-size:18px; font-weight:bold; text-align:center;'>Figure 4: Poverty Distribution by Region</div>"), 
    position = "topleft"
  )
```

Comparing with the total poverty percentage per state, I observe similar density distribution across the country. This can be explained that higher poverty rates leads to lower healthcare access, which could then lead to higher disability rates. States with better healthcare infrastructure and higher income levels tend to have lower disability rates.

```{r echo=FALSE}
# Define disability types
disability_types <- c("SelfCareDisability", "HearingDisability", "VisionDisability",
                      "IndependentLivingDisability", "MobilityDisability", "CognitiveDisability",
                      "AnyDisability")

# Initialize an empty list to store ANOVA results
anova_summary_list <- list()

# Perform ANOVA and extract relevant statistics
for (disability in disability_types) {
  aov_result <- aov(df_final[[disability]] ~ df_final$State)
  summary_result <- summary(aov_result)[[1]]  # Extract summary table
  
  # Convert to dataframe and rename columns properly
  anova_df <- as.data.frame(summary_result) %>%
    rownames_to_column(var = "Term") %>%  # Preserve row names as a column
    filter(Term == "df_final$State") %>%  # Keep only 'df_final$State' row
    mutate(Disability = disability) %>%   # Add disability type column
    select(Disability, Df, `Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)`)  # Keep only required columns
  
  # Store the formatted table
  anova_summary_list[[disability]] <- anova_df
}

# Combine results for all disabilities
anova_summary_table <- bind_rows(anova_summary_list)

# Format p-values and highlight significance levels
anova_summary_table <- anova_summary_table %>%
  mutate(
    `Pr(>F)` = ifelse(`Pr(>F)` < 0.001, paste0(sprintf("%.3f", `Pr(>F)`), "***"),
                      ifelse(`Pr(>F)` < 0.01, paste0(sprintf("%.3f", `Pr(>F)`), "**"),
                             ifelse(`Pr(>F)` < 0.05, paste0(sprintf("%.3f", `Pr(>F)`), "*"),
                                    sprintf("%.3f", `Pr(>F)`))))
  )

# Pretty table using kableExtra
kable(anova_summary_table, caption = "Table 1: ANOVA Results for Disability Proportions Across States",
      digits = 3, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE) %>%  # Bold disability type column
  row_spec(0, bold = TRUE, align = "c")
```

To further investigate whether these observed geographic variations in disability prevalence are statistically significant, I conducted an ANOVA (Analysis of Variance) test. From Table 1, the ANOVA results showed p-values are all < 2e-16 (very close to 0). This means there is a strong statistical evidence that disability proportions significantly differ across states. Since p < 0.05, I reject the null hypothesis and conclude that disability proportions differ meaningfully across states beyond what could be attributed to random variation. By analyzing different disability types (e.g., mobility, cognitive, self-care), I can assess which forms of disability exhibit the most substantial regional disparities and how these differences align with socioeconomic patterns. The following table presents the ANOVA results, highlighting statistically significant variations across states.

```{r include=FALSE}
# Normalize data
df_cluster <- df_final %>% select(all_of(disability_types))
df_cluster_scaled <- scale(df_cluster)

# Apply K-means clustering
set.seed(123)
kmeans_model <- kmeans(df_cluster_scaled, centers = 3)

# Add cluster labels
df_final$Cluster <- as.factor(kmeans_model$cluster)

# Visualize clusters
ggplot(df_final, aes(x = MobilityDisability, y = SelfCareDisability, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "Clustering States by Disability Patterns", x = "Mobility Disability Proportion", y = "Self-Care Disability Proportion") +
  theme_minimal()
```
```{r echo=FALSE}
# Define disability types
disability_types <- c("SelfCareDisability", "HearingDisability", "VisionDisability",
                      "IndependentLivingDisability", "MobilityDisability", "CognitiveDisability",
                      "AnyDisability")

# Define independent variables (predictors)
predictors <- c("DisabilityPrevalence", "UninsuredRate", "PovertyTotal", "MedicaidInsuranceCoverage",
                "EmployerInsuranceCoverage", "MedicareInsuranceCoverage", "Total.Hospitals",
                "Total.Hospital.Beds", "TotalMedicareEnrollees", "DisabilityMedicaidSpending")

# Initialize an empty list to store results
summary_list <- list()

# Loop through each disability type and fit a linear regression model
for (target in disability_types) {
  # Create formula dynamically
  formula <- as.formula(paste(target, "~", paste(predictors, collapse = " + ")))

  # Fit the linear model
  model <- lm(formula, data = df_final)

  # Extract coefficients and p-values
  model_summary <- summary(model)$coefficients %>%
    as.data.frame() %>%
    select(Estimate, `Pr(>|t|)`) %>%
    rename(`p-value` = `Pr(>|t|)`) %>%
    mutate(
      Predictor = rownames(.),
      Estimate = ifelse(`p-value` < 0.05, paste0(round(Estimate, 2), "*"), round(Estimate, 2)) # Add * for significance
    ) %>%
    select(Predictor, Estimate)  # Keep only required columns

  # Store results with the disability type as a column
  summary_list[[target]] <- model_summary
}

# Combine all results into a single dataframe (long format)
summary_table <- bind_rows(summary_list, .id = "Disability Type")

# Reshape to **wide format** (Predictors as rows, Disabilities as columns)
summary_wide <- summary_table %>%
  pivot_wider(names_from = "Disability Type", values_from = Estimate)

# Remove "Disability" from column names
colnames(summary_wide) <- gsub("Disability", "", colnames(summary_wide))

colnames(summary_wide)[1] <- "Predictor | Disability Types"

# Generate the table with correct formatting
kable(summary_wide, caption = "Table 2: Comparison of Estimates Across Disabilities (Significant values marked with *)", 
      digits = 2, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE) %>%  # Make predictors bold
  row_spec(0, bold = TRUE, align = "c")  # Bold and center-align the header row
```
The regression results in Table 2 highlight significant associations between various socioeconomic factors and disability prevalence across different disability types. Statistically significant coefficients (denoted by \*) identify the most influential predictors. Poverty levels show a strong positive correlation with disability prevalence, particularly for Mobility Disability, Cognitive Disability, and Any Disability. This suggests that individuals in higher poverty brackets are more likely to report disabilities, potentially due to limited access to healthcare and preventive services, greater exposure to environmental and occupational risk factors, and an increased likelihood of chronic health conditions. Medicare and Medicaid coverage also play a crucial role in disability prevalence. Medicare Insurance Coverage is significantly associated with higher prevalence of Vision Disability, Independent Living Disability, Mobility Disability, Cognitive Disability, and Any Disability. Given that Medicare primarily serves individuals aged 65 and older, as well as certain disabled individuals, this trend aligns with expectations that older populations and those with disabilities are more likely to be enrolled in Medicare. Similarly, Medicaid Insurance Coverage exhibits a significant positive relationship with Cognitive Disability and Self-Care Disability, reinforcing the connection between lower-income populations, Medicaid utilization, and higher disability rates. Employer-sponsored insurance coverage, on the other hand, is negatively associated with several disability types, including Hearing Disability, Independent Living Disability, Mobility Disability, and Any Disability. This suggests that individuals with employer-sponsored insurance are less likely to report disabilities, potentially due to better access to healthcare services, workplace accommodations, and overall improved socioeconomic conditions. While some variables, such as the uninsured rate, exhibit sporadic significance, their overall effect sizes are minimal, suggesting a limited impact on disability prevalence. Hospital-related factors, including Total Hospitals, Hospital Beds, and Total Medicare Enrollees, display weak or mixed effects, indicating that while healthcare infrastructure plays a role, it may not be a primary determinant of disability rates. Additionally, Disability Medicaid Spending is statistically significant in select cases but with small coefficient values, suggesting it is not a major explanatory factor in disability prevalence. These findings carry important policy implications. Given the strong link between poverty and disability, expanding access to preventive healthcare, rehabilitation services, and disability accommodations could help mitigate disparities. The significant association between Medicaid coverage and cognitive disabilities suggests that policies aimed at expanding mental health services could be particularly beneficial. Furthermore, states with aging populations, such as Florida, may see higher rates of hearing and vision-related disabilities, necessitating age-friendly healthcare policies and accessibility improvements.

### Summary

So far, my analysis has revealed strong associations between socioeconomic factors and disability prevalence across U.S. states. My findings support the hypothesis that states with higher poverty rates, lower healthcare coverage, and greater economic inequality tend to have higher disability prevalence. Specifically, poverty is positively correlated with Mobility Disability, Cognitive Disability, and Any Disability, while Medicaid and Medicare coverage are also significant predictors of disability prevalence. On the other hand, employer-sponsored insurance is negatively associated with disability rates, suggesting that access to workplace benefits may help mitigate disability risk.

Preliminary geospatial analysis using Leaflet maps shows that disability prevalence is higher in Southern states and lower in Western and Northeastern states. This aligns with socioeconomic disparities, where states with lower income levels and weaker healthcare infrastructure tend to experience higher disability rates. Additionally, correlation analysis indicates that higher Medicaid spending is linked to lower uninsured rates, reinforcing the role of public health programs in supporting individuals with disabilities.  

To further strengthen my findings and provide deeper insights, I plan to take the following steps. I plan to do clustering analysis (K-Means or Hierarchical Clustering), where I can identify clusters of states with similar disability prevalence patterns and determine whether these clusters align with socioeconomic characteristics. This will reveal distinct clusters of states with similar disability trends, helping policymakers target region-specific interventions. I would also like to try spatial analysis and explore geographic dependencies to test whether disability prevalence exhibits spatial dependencies and regional spillover effects, so I can determine whether policy interventions should be regionally tailored due to geographic effects. Finally, I can compare findings with existing healthcare policies and economic programs and suggest potential policy interventions based on statistical and spatial analysis results.  

By integrating statistical modeling, clustering, and spatial analysis, the final project will provide a comprehensive, data-driven understanding of how socioeconomic and demographic factors impact disability prevalence across the U.S.